---
title: "Comparison methods"
author: "Sophia Müller-Dott"
date: '2022-09-14'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
library(org.Hs.eg.db)
library(KSEAapp)
library(tidyverse)
source("code/method_functions.R")
source("code/support_functions.R")
```

## Introduction

Here we will compare the kinase actvities of each method. From the methods in the review these will be considered in the compatison:

Considered:
- IKAP
- KARP
- KSEA App
- PTM-SEA
- RoKAI
- KEA3 (only for Venn diagramm)

Not considered:
- CLUE: time-series
- KinasePA: multi-condition studies
- INKA: code only available for MaxQuant files

We tried to run the methods with their default settings. The methods were run for three CPTAC studies (ccrcc, luad, ucec)

## Kinase activity estimation
For the methods we used comparison between tumor samples and normal samples from three CPTAC studies (ccrcc, luad, ucec).

```{r}
## Load data ---------------------------
CPTAC_data <- list(ccrcc = read.csv("data/CPTAC_estefani_clean_zip/ccrcc/ccrcc_phospho_tvalues.csv"),
                   luad = read.csv("data/CPTAC_estefani_clean_zip/luad/luad_phospho_tvalues.csv"),
                   ucec = read.csv("data/CPTAC_estefani_clean_zip/ucec/ucec_phospho_tvalues.csv"))

CPTAC_data <- map(CPTAC_data, function(CPTAC_cohort){
  cbind(protein_symbol =  map_chr(str_split(CPTAC_cohort$ID, "_"), function(x) x[1]),
        pps =  map_chr(str_split(CPTAC_cohort$ID, "_"), function(x) x[2]),
        CPTAC_cohort)
})

## Add uniprot ID, filter NAs in t-values, filter peptides with multipls pps
CPTAC_data_filtered <- map(CPTAC_data, function(CPTAC_cohort){
  mapped <- mapIds(org.Hs.eg.db,
                   keys = CPTAC_cohort$protein_symbol,
                   keytype = "SYMBOL",
                   column = c("UNIPROT"))

  CPTAC_cohort <- CPTAC_cohort %>%
    add_column(protein_uniprot = mapped, .after = "protein_symbol") %>%
    filter(rowSums(is.na(.)) == 0) %>%
    filter(str_count(pps, "[A-Z]") == 1)

  CPTAC_cohort %>% add_column(uniprot_ID = paste0(CPTAC_cohort$protein_uniprot, "_", CPTAC_cohort$pps), .after = "ID")

})

```


### IKAP
t values
input preparation
```{r}
# Network
PPS.plus <- read_tsv(file = "data/PPSp_20220830", skip = 2)
IKAP.network <- PPS.plus %>% 
  dplyr::filter(KIN_ORGANISM == "human" & SUB_ORGANISM == "human") %>%
  dplyr::select(KINASE, SUB_GENE, SUB_MOD_RSD) %>%
  dplyr::mutate(PSP = paste(SUB_GENE, SUB_MOD_RSD, sep = "_")) %>%
  dplyr::rename("GENE" = "KINASE") %>%
  dplyr::select(GENE, PSP)

# Samples
map(names(CPTAC_data_filtered), function(cohort){
 cohort_data <- CPTAC_data_filtered[[cohort]]
 
  IKAP_data <- data.frame(str_split_fixed(cohort_data$ID, "_", 2)[,1]) %>%
    add_column(peptide =cohort_data$ID,
      logFC =cohort_data$NATvsTUM_t)

  colnames(IKAP_data)[1] <- "protein"
  write.csv(IKAP_data, paste0("code/IKAP/IKAP_input_CPTAC_",cohort, ".csv"), row.names = F)
})

write.csv(IKAP.network, paste0("code/IKAP/phosphositeplus.csv"), row.names = F)
```

IKAP was run in MATLAB and the results were processed in R 
```{r}
IKAP_act_files <- list.files("code/IKAP", pattern = "kinase", full.names = T)
IKAP_names <- str_remove(str_remove(IKAP_act_files, "code/IKAP/kinase_act_"), ".csv")
IKAP_names <- str_remove(IKAP_names, "code/IKAP/kinase_")

IKAP_act <- map(unique(IKAP_names), function(cohort){
  files <- list.files("code/IKAP", pattern = cohort, full.names = T)
  act_df <- cbind(read_csv(files[3], col_names = F), read_csv(files[2], col_names = F))
  names(act_df) <- c("Kinase", "Score")
  act_df %>%
    arrange(desc(abs(Score)))
})

names(IKAP_act) <- unique(IKAP_names)

head(IKAP_act$ccrcc)
```

### KARP
t values
input preparation
```{r}
# Network
PPS.plus.KARP <- read_tsv(file = "data/PPSp_20220830", skip = 2)  %>% 
  dplyr::filter(KIN_ORGANISM == "human" & SUB_ORGANISM == "human") %>%
  dplyr::select(KINASE, SUB_GENE, SUB_MOD_RSD) %>%
  dplyr::mutate(TARGET = paste(SUB_GENE, SUB_MOD_RSD, sep = "_")) %>%
  dplyr::rename("GENE" = "KINASE") %>%
  dplyr::select(GENE, TARGET)

# Samples
KARP.input <- map(names(CPTAC_data_filtered), function(cohort){
 cohort_data <- CPTAC_data_filtered[[cohort]]
 
 cohort_data <- cohort_data %>% 
   dplyr::select(ID, NATvsTUM_t)
})
```

```{r}
KARP_act <- map(KARP.input, function(sample){
  act_df <- runKARP(t.values = sample, PKN = PPS.plus.KARP, min.targets = 1)
  names(act_df) <- c("Kinase", "Score")
  act_df %>%
    arrange(desc(abs(Score)))})

names(KARP_act) <- names(CPTAC_data_filtered)

head(KARP_act$ccrcc)
```

### KSEA App
t values
input preparation
```{r}
KSEA.input <- map(names(CPTAC_data_filtered), function(cohort){
 cohort_data <- CPTAC_data_filtered[[cohort]]
  KSEA_data <- data.frame(str_split_fixed(cohort_data$ID, "_", 2)) %>%
    tibble::add_column(p =cohort_data$NATvsTUM_adj.P.Val,
               FC = (2^cohort_data$NATvsTUM_t),
               Protein = str_split_fixed(cohort_data$uniprot_ID, "_", 2)[,1],
               Peptide = "NULL") %>%
    dplyr::rename("Gene" = "X1", "Residue.Both" = "X2")

  KSEA_data <- KSEA_data[c(5,1,6,2,3,4)]
})
```

```{r}
KSEA_act <- map(KSEA.input, function(sample){
  KSEA.network <- KSEAapp::KSData

  KSEA_scores <- KSEA.Scores(KSData = KSEA.network,
                             PX = sample,
                             NetworKIN = F,
                             NetworKIN.cutoff = 0)

  KSEA_scores %>%
    dplyr::select(Kinase.Gene, z.score) %>%
    dplyr::rename(c("Kinase" = "Kinase.Gene", "Score" = "z.score")) %>%
    arrange(desc(abs(Score)))
    
})

names(KSEA_act) <- names(CPTAC_data_filtered)

head(KSEA_act$ccrcc)
```
```{r}
KSEA.input <- map(names(CPTAC_data_filtered), function(cohort){
 cohort_data <- CPTAC_data_filtered[[cohort]]
 cohort_data %>%
   dplyr::select(ID, NATvsTUM_t) %>%
   column_to_rownames("ID")
})
```

```{r}
SEA_act <- map(KSEA.input, function(sample){
  
  KSEA.network <- KSEAapp::KSData
  network <- data.frame(source = KSEA.network$GENE,
                        target = paste0(KSEA.network$SUB_GENE, "_", KSEA.network$SUB_MOD_RSD),
                        mor = 1)

  KSEA_scores <- run_zscore_KSEA(network = network,
                                mat = sample)
  })

names(SEA_act) <- names(CPTAC_data_filtered)

head(SEA_act$ccrcc)

full_join(SEA_act$ccrcc, KSEA_act$ccrcc %>% dplyr::rename("source" = "Kinase"), by = c("source"))
```

### PTM-SEA
t values
input preparation
```{r}
map(names(CPTAC_data_filtered), function(cohort){
 cohort_data <- CPTAC_data_filtered[[cohort]]
 cohort_data <- cohort_data %>%
    dplyr::mutate(rid = paste0(str_replace(uniprot_ID, "_", ";"), "-p")) %>%
    dplyr::select(c("rid", "NATvsTUM_t")) %>%
    dplyr::rename("t" = "NATvsTUM_t") %>%
    column_to_rownames("rid")
 cohort_data_m <- as.matrix(cohort_data)

 cohort_data_gct <- new("GCT", mat=cohort_data_m)
 write_gct(cohort_data_gct, paste0("code/PTM-SEA/PTM-SEA_input_",cohort))
})

```
```{r}
PTM_act_files <- list.files("code/PTM-SEA/2022-09-14", pattern = "ssGSEA-scores.gct", full.names = T, recursive = TRUE)
PTM_names <- map_chr(str_split(PTM_act_files, "_"), function(x) x[3])
PTM_act <- map(PTM_act_files, function(x){
  act <- read.delim(file = x, skip=2)
  PTM <- act[str_detect(act$id, "KINASE"), c("id", "t")] %>%
    dplyr::rename(c("Kinase" = "id", "Score" = "t"))
  PTM$Kinase <- map_chr(str_split(PTM$Kinase, "_"), function(x) x[2])
  PTM$Kinase <-  map_chr(str_split(PTM$Kinase, "/"), function(x) x[length(x)])
  PTM %>%
    arrange(desc(abs(Score)))
})
names(PTM_act) <- PTM_names

head(PTM_act$ccrcc)
```


### RoKAI
t value
input preparation
```{r}
map(names(CPTAC_data_filtered), function(cohort){
 cohort_data <- CPTAC_data_filtered[[cohort]]
  rokai_data <- data.frame(str_split_fixed(cohort_data$uniprot_ID, "_", 2)) %>%
    add_column(Quantification = cohort_data$NATvsTUM_t)

  colnames(rokai_data) <- c("Protein", "Position", "Quantification")

  write.csv(rokai_data, paste0("code/RoKAI/rokai_input_",cohort, ".csv"))
})
```
RoKAI kinase activity estimation was run using the web application https://rokai.io/
```{r}
rokai_act_files <- list.files("code/rokai", pattern = "kinase_table", full.names = T)
rokai_names <- str_remove(str_remove(rokai_act_files, "code/rokai/kinase_table_"), ".csv")
rokai_act <- map(rokai_act_files, function(file){
  res <- read_csv(file)
  res %>% 
    dplyr::select(Gene, ZScore) %>%
    dplyr::rename(c("Kinase" = "Gene", "Score" = ZScore)) %>%
    arrange(desc(abs(Score)))
  }
  )
names(rokai_act) <- rokai_names

head(rokai_act$ccrcc)
```


### KEA3
```{r}
map(names(CPTAC_data_filtered), function(cohort){
 cohort_data <- CPTAC_data_filtered[[cohort]]

  KEA3_data <-cohort_data %>%
    filter(NATvsTUM_adj.P.Val <= 0.05) %>%
    filter(abs(NATvsTUM_logFC) >= 2)

  KEA3_data <- unique(KEA3_data$protein_symbol)

  write.table(KEA3_data,  paste0("code/KEA3/KEA3_input_", cohort, ".txt"), row.names = F, col.names = F,  quote = FALSE)

})
```
We also used the web application of KEA3 (https://maayanlab.cloud/kea3/)
```{r}
KEA3_act_files <- list.files("code/KEA3", pattern = "mean_rank", full.names = T)
KEA3_names <- str_remove(str_remove(KEA3_act_files, "code/KEA3/mean_rank_"), ".tsv")
KEA3_act <- map(KEA3_act_files, function(file){
  res <- read_tsv(file)
  res$Protein <- str_remove(res$Protein, "\\*")
  res %>% 
    dplyr::select(Rank, Protein) %>%
    dplyr::rename("Kinase" = "Protein", "Score" = "Rank")
})

names(KEA3_act) <- KEA3_names

head(KEA3_act$ccrcc)
```


## Comparison
### Correlation
```{r}
cohort <- c("ccrcc", "luad", "ucec")
results <- map(cohort, function(i){
           list(IKAP = IKAP_act[[i]],
                KARP = KARP_act[[i]],
                KSEA = KSEA_act[[i]],
                PTM = PTM_act[[i]],
                RoKAI = rokai_act[[i]],
                KEA3 = KEA3_act[[i]])
})
names(results) <- cohort

results_df <- map(results, function(res){
  sel <- names(res)[! names(res) == "KEA3"]
  res_df <- res[sel] %>% 
    purrr::reduce(full_join, by = "Kinase") %>%
    filter(!base::duplicated(Kinase)) %>%
    column_to_rownames("Kinase") 
  colnames(res_df) <- sel
  
  res_df
})

results_df <- append(results_df, list(all = bind_rows(results_df)))


corr_df <- map(results_df, function(act_df){
  corr_matrix <- matrix(nrow = ncol(act_df),
                        ncol = ncol(act_df))
  
  for (i in 1:ncol(act_df)) {
    for (j in 1:ncol(act_df)) {
      pearson.cor <- stats::cor.test(act_df[,i], act_df[,j])
      corr_matrix[i,j] <- pearson.cor$estimate
    }
  }
  corr_df <- as.data.frame(corr_matrix)
  colnames(corr_df) <- colnames(act_df)
  rownames(corr_df) <- colnames(act_df)
  
  corr_df
})

map(corr_df, function(corr_cohort){
  get_mat_plot(corr_cohort, main = 'Pearson correlation', palette = "Greens")
})

mean_corr_matrix <- matrix(ncol = ncol(corr_df$ccrcc),
                           nrow = nrow(corr_df$ccrcc))

for (i in 1:ncol(corr_df$ccrcc)){
  for (j in 1:nrow(corr_df$ccrcc)){
    mean_corr_matrix[j,i] <-  mean(c(corr_df$ccrcc[j,i],
                                  corr_df$luad[j,i],
                                  corr_df$ucec[j,i]))

  }

}

colnames(mean_corr_matrix) <- colnames(corr_df$ccrcc)
rownames(mean_corr_matrix) <- rownames(corr_df$ccrcc)
```


```{r}
mat <- corr_df$all
# Get order of clustered methods
cor_heat <- pheatmap::pheatmap(mat, cluster_rows = T,
                              cluster_cols = T, silent=T)
idxRows <- cor_heat$tree_row$order
idxCols <- cor_heat$tree_col$order
mat <- mat[idxRows,idxCols]
  
# Plot
paletteLength <- 100
color <- colorRampPalette(c("whitesmoke","#AD4747"))(paletteLength)

heat_p <- ComplexHeatmap::Heatmap(mat, col = color, 
                                  show_column_dend = FALSE, 
                                  row_names_gp = grid::gpar(fontsize = 10), 
                                  column_names_gp = grid::gpar(fontsize = 10),
                                  row_dend_width = unit(0.4, "inch"), show_heatmap_legend = FALSE
                                  )

col_fun <- circlize::colorRamp2(c(0,  1), c("whitesmoke","#AD4747"))
lgd <- ComplexHeatmap::Legend(col_fun = col_fun, title = "PCC", legend_height = unit(1.1, "inch"), 
                              title_gp = gpar(fontsize = 8, fontface = "plain"), labels_gp = gpar(fontsize = 8))

pdf("output/figures/legend_heatmap.pdf", width = 0.8, height = 1.7)
draw(lgd)
dev.off()

pdf("output/figures/correlation_heatmap.pdf", width = 2.0, height = 1.7)
heat_p
dev.off()
```
### Scatter plots

```{r}
results_long_list <- map(names(results_df), function(x){
  results_df[[x]] %>% 
  rownames_to_column("Kinase") %>%
    add_column(dataset = x)
})
results_long <- do.call("rbind", results_long_list) %>%
  filter(!dataset == "all")

methods <- c("KSEA", "RoKAI", "PTM", "IKAP", "KARP")
combinations <- apply(combn(methods,2),2,paste,collapse='_')

scatter_p <- map(combinations, function(method_comp){
  meth_id <- str_split(method_comp, "_") %>% unlist()
  
  df_plot <- results_long[c("Kinase", meth_id, "dataset")]
  colnames(df_plot) <- c("Kinase", "Method1", "Method2", "dataset")
  
  ggplot(df_plot) + 
  geom_point(aes(x=Method1, y=Method2, color=dataset), size=0.5) +
  geom_smooth(aes(x=Method1, y=Method2), method=lm, fullrange=TRUE, color = "darkgrey", fill = "lightgrey", size=0.3) + 
  scale_color_manual(values=c('#47ADAD','#477AAD', '#ADAD47'), name = "") +
  theme_minimal() + xlab(meth_id[1]) + ylab(meth_id[2]) + 
  theme(text = element_text(size = 10)) 
  
})

leg <- ggpubr::get_legend(scatter_p[[1]])
```
```{r}
pdf("output/figures/scatter_plots.pdf", width = 6.25, height = 7.5)
do.call(ggpubr::ggarrange, c(append(list(NULL, ggpubr::as_ggplot(leg)), scatter_p),  ncol = 3, nrow = 4, common.legend = T, legend = "bottom"))
do.call(ggpubr::ggarrange, c(append(list(NULL, NULL), scatter_p),  ncol = 3, nrow = 4, common.legend = T, legend = "bottom"))
dev.off()

```




### Top kinases
Here we will select top n kinases based on the absolute score and check how much they overlap between the methods
```{r}
n <- 10

top_kinases <- map(results, function(activities){
  map(activities, function(method){
    method %>% 
      dplyr::slice(1:n) %>%
      dplyr::pull(Kinase) %>%
      as.character()
  })
})

venn_p <- map(top_kinases, function(x){
  m <- data.frame(elements = unlist(x), sets = rep(names(x), each = n))
  v <- venneuler::venneuler(m)
})

for(cptac in names(venn_p)) {
  venn <- venn_p[[cptac]]
  venn$labels <- rep("", length(venn$labels))
  pdf(paste0("output/figures/venn_", cptac,".pdf"), width = 2.1, height = 2.1)
  par(mar=c(0.5,0.5,0.5, 0.5))
  plot(venn)
  for (i in 1:nrow(venn$centers)) {
    text(venn$centers[i,1], venn$centers[i,2], rownames(venn$centers)[i], cex = 0.85)
  }
  dev.off()
}

```

